---
title: "Introduction to edfun"
date: "`r Sys.Date()`"
author: "Tal Galili"
output:
  html_document:
    self_contained: yes
    toc: true    
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Introduction to edfun}
-->


```{r, echo = FALSE, message = FALSE}
library(edfun)
library(knitr)
knitr::opts_chunk$set(
   # cache = TRUE,
   dpi = 60,
  comment = "#>",
  tidy = FALSE)

# http://stackoverflow.com/questions/24091735/why-pandoc-does-not-retrieve-the-image-file
# < ! -- rmarkdown v1 -->

```


**Author**: [Tal Galili](http://www.r-statistics.com/) ( Tal.Galili@gmail.com )



Introduction
============


As mentioned in [CRAN Task View: Probability Distributions](https://cran.r-project.org/web/views/Distributions.html)

> Empirical distribution : Base R provides functions for univariate analysis: (1) the empirical density (see density()), (2) the empirical cumulative distribution function (see ecdf()), (3) the empirical quantile (see quantile()) and (4) random sampling (see sample()).

This package aims to easily wrap these into a single function `edfun` (short for Empirical Distribution FUNctions). Also, since quantile is generally a slow function to perform, the default for creating a quantile function (inverse-CDF) is by approximating the function of predicting the data values (x) from their quantiles (CDF). This is done using the `approxfun` function. It takes a bit longer to create qfun, but it is MUCH faster to run than quantile (and is thus much better for simulations). Special care is taken for dealing with the support of the distribution (if it is known upfront).

The added speed allows to use these functions to run simulation studies for unusual distributions.


Installation
============

To install the stable version on CRAN:

```{r, eval = FALSE}
# install.packages('edfun') # not on CRAN yet
```

To install the GitHub version:

```{r, eval = FALSE}
# You'll need devtools
if (!require(devtools)) install.packages("devtools");

devtools::install_github('talgalili/edfun')
```


And then you may load the package using:

```{r}
library("edfun")
```

Usage
======

Basic usage - the normal distribution
--------------------------------------

First load the library:

```{r}
library(edfun)
```



```{r}

set.seed(123)
x <- rnorm(1000)
x_dist <- edfun(x)

f <- x_dist$dfun
curve(f, -2,2)

f <- x_dist$pfun
curve(f, -2,2)

f <- x_dist$qfun
curve(f, 0,1)

new_x <- x_dist$rfun(1000)
hist(new_x)
```

The same can be done if we had the density of the distribution. In this case, `x` should be a sequence of values for which we wish to pre-compute the quantiles (inv-CDF).


```{r}

x <- seq(-4,4, length.out = 1e3)
x_dist <- edfun(x, dfun = dnorm)

f <- x_dist$dfun
curve(f, -2,2)

f <- x_dist$pfun
curve(f, -2,2)

f <- x_dist$qfun
curve(f, 0,1)

new_x <- x_dist$rfun(1000)
hist(new_x)
```



If it is known, we can specifically define the support:

```{r}

x <- seq(-4,4, length.out = 1e3)
x_dist_no_support <- edfun(x, dfun = dnorm)
x_dist_known_support <- edfun(x, dfun = dnorm, support = c(-Inf, Inf))

x_dist_no_support$qfun(0)
x_dist_known_support$qfun(0)

```




Timing is basically the same when using the functions relying on density (i.e.: `dfun`) or on a sample only. But the accuracy of using the density is much better.

```{r, eval = FALSE}
 set.seed(2016-08-18)
  x_simu <- rnorm(1e5)
  x_funs_simu <- edfun(x_simu, support = c(-Inf, Inf))
  
  
  x <- seq(-4,4, length.out = 1e5)
  x_funs <- edfun(x, dfun = dnorm, support = c(-Inf, Inf))
  x_funs$qfun(0) # -Inf
  
  # precision - qfun
  q_to_test <- seq(0.001,.999, length.out = 100)
  edfun_out <- x_funs$qfun(q_to_test) # -4
  edfun_simu_out <- x_funs_simu$qfun(q_to_test) # -4
  real_out <- qnorm(q_to_test)
  
  mean(abs(edfun_out-real_out))
  mean(abs(edfun_simu_out-real_out)) # quite terrible compared to when knowing dfun
  
  library(microbenchmark)
  microbenchmark(dfun_known = x_funs$qfun(q_to_test),
                 dfun_NOT_known = x_funs_simu$qfun(q_to_test))
  # same time for the q function!
  
  
  # precision - pfun
  q_to_test <- seq(-3,3, length.out = 100)
  edfun_out <- x_funs$pfun(q_to_test) # -4
  edfun_simu_out <- x_funs_simu$pfun(q_to_test) # -4
  real_out <- pnorm(q_to_test)
  
  mean(abs(edfun_out-real_out))
  mean(abs(edfun_simu_out-real_out)) # quite terrible compared to when knowing dfun
  
  library(microbenchmark)
  microbenchmark(dfun_known = x_funs$pfun(q_to_test),
                 dfun_NOT_known = x_funs_simu$pfun(q_to_test))
  # same time for the p function!

  
  # timing for the rfun
  library(microbenchmark)
  microbenchmark(dfun_known = x_funs$rfun(100),
                 dfun_NOT_known = x_funs_simu$rfun(100))
  # this rfun is slower...  
```


The double exponential distribution
--------------------------------------

First load the library:

```{r}
library(edfun)
```


```{r}

# credit: library(smoothmest)
ddoublex <- function (x, mu = 0, lambda = 1) {
  exp(-abs(x - mu)/lambda)/(2 * lambda)  
}

curve(ddoublex, -4,4) # the peak in 0 should go to Inf, but it doesn't because of the limits of `curve`

x <- seq(-4,4, length.out = 1e3)
x_dist <- edfun(x, dfun = ddoublex)

f <- x_dist$dfun
curve(f, -4,4)

f <- x_dist$pfun
curve(f, -4,4)

f <- x_dist$qfun
curve(f, 0,1)

new_x <- x_dist$rfun(1000)
hist(new_x)
```





A mixture of two normal distributions
--------------------------------------

First load the library:

```{r}
library(edfun)
```


```{r}

# http://stackoverflow.com/questions/20452650/writing-a-bimodal-normal-distribution-function-in-r?rq=1
# random sample:
# mixtools::rnormmix # random sample of a mixture of distributions
# http://stats.stackexchange.com/questions/70855/generating-random-variables-from-a-mixture-of-normal-distributions
# nor1mix::rnorMix
# https://en.wikipedia.org/wiki/Mixture_distribution
# density:
dmixnorm <- function(x, p1 = 0.5, m1=0, m2=0, s1=1, s2=1) {
  p1 * dnorm(x, m1, s1) + (1-p1) * dnorm(x, m2, s2)
}
# a convex mixture of densities is a density:
#  https://en.wikipedia.org/wiki/Mixture_distribution#Properties

# yay - it works.
dmixnorm_1 <- function(x) dmixnorm(x, .75, 0,5,  1,1)
curve(dmixnorm_1, -4,9)


x <- seq(-4,9, length.out = 1e3)
x_dist <- edfun(x, dfun = dmixnorm_1)

f <- x_dist$dfun
curve(f, -4,9)

f <- x_dist$pfun
curve(f, -4,9)

f <- x_dist$qfun
curve(f, 0,1)

new_x <- x_dist$rfun(1000)
hist(new_x)
```





Contact
============


You are welcome to:

* submit suggestions and bug-reports at: <https://github.com/talgalili/edfun/issues>
* send a pull request on: <https://github.com/talgalili/edfun/>
* compose a friendly e-mail to: <tal.galili@gmail.com>


Latest news
===============

You can see the most recent changes to the package in the [NEWS.md file](https://github.com/talgalili/edfun/blob/master/NEWS.md)





Session info
=============

```{r, cache=FALSE}
sessionInfo()
```
